---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Fiona Wang"
date: "2025-02-25"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox)
library(mosaicData) 
library(scales)
library(forcats)
```

Load data
```{r message=FALSE}
staff <- read_csv("data/instructional-staff.csv")
#reshape from wide format to long format
staff_long <- staff %>% 
  pivot_longer(cols = -faculty_type, names_to = "year") %>% 
  mutate(value = as.numeric(value))
staff_long
```

Plot the graph
```{r}
staff_long %>% 
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line()
```

### Exercise 1
```{r fig.height=7, fig.width=10}
staff_long %>% 
  ggplot(aes(
    x = year,
    y = value / 100,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  labs(title = "Faculty Hiring Trends: Percentage of Each Type Hired Per Year",
       subtitle = "1975 - 2011",
       x = "Year",
       y = NULL,
       color = "Faculty Type") + 
  scale_y_continuous(labels = percent_format())
```

### Exercise 2
To tell the story that part-time faculty hiring has been increasing while other faculty types have been decreasing in hiring percentages, I think changing the colors of the lines, as well as the line types would make this contrast stands out more. 
```{r fig.height=7, fig.width=10}
staff_long %>% 
  ggplot(aes(
    x = year,
    y = value / 100,
    group = faculty_type,
    color = faculty_type,
    linetype = faculty_type
  )) +
  geom_line() +
  labs(title = "Faculty Hiring Trends: Percentage of Each Type Hired Per Year",
       subtitle = "1975 - 2011",
       x = "Year",
       y = NULL,
       color = "Faculty Type") + 
  scale_y_continuous(labels = percent_format()) +
  scale_color_manual(values = c(
    "Full-Time Tenured Faculty" = "#2b8cbe",
    "Full-Time Tenure-Track Faculty" = "#2b8cbe",
    "Full-Time Non-Tenure-Track Faculty" = "#2b8cbe",
    "Graduate Student Employees" = "#2b8cbe",
    "Part-Time Faculty" = "#de2d26"
  )
  ) +
  scale_linetype_manual(values = c(
    "Full-Time Tenured Faculty" = "longdash",
    "Full-Time Tenure-Track Faculty" = "longdash",
    "Full-Time Non-Tenure-Track Faculty" = "longdash",
    "Graduate Student Employees" = "longdash",
    "Part-Time Faculty" = "solid"
  ))
# I don't know if there is a more concise way to write the code.
```

### Exercise 3
```{r message=FALSE}
fisheries <- read_csv("data/fisheries.csv")
```

How to improve the graph?
The first density plot is not very legible. All the countries, except China, have very low frequencies. For the pie chart, it looks pretty fancy, it's just that I can't tell which part is which country. 
```{r fig.height=7, fig.width=14}
fisheries_aqua <- fisheries %>% 
  arrange(desc(aquaculture)) %>% 
  slice(c(1:16))
fisheries_cap <- fisheries %>% 
  arrange(desc(capture)) %>% 
  slice(c(1:16))
fisheries_total <- fisheries %>% 
  arrange(desc(total)) %>% 
  slice(c(1:16))
fisheries_total %>% 
  arrange(desc(total)) %>% 
  ggplot(aes(x = reorder(country, -total), y = total, fill = "pink")) + 
  geom_col(position = "dodge") + 
  geom_text(aes(label = comma(total))) +
  scale_y_continuous(labels = comma) +
  scale_x_discrete(labels = ~str_wrap(.x, width = 10)) + 
  labs(title = "Tonnage of fish captured and farmed in total for each country",
       x = "Country",
       y = "Tonnage") +
  theme(text = element_text(size = 12.5))
```

This graph here only represents the total tonnage (capture + aquaculture).   
What I changed:    
1. I used bar plot instead of density/line plot. This is because a trend between countries doesn't make a lot of sense.      
2. I labeled the exact values on each bar. Because China has a very high value, other countries have shorter bars in the graph. It's not easy to make sense of the value of those countries.      
3. Added labels so that it's clear what this graph is about.     
4. Width of the graph so that it's more readable.         

One thing that could be changed: represents capture, aquaculture, and total in the same graph. 

Let's try doing that. This will need to include what we just learned: pivot_longer.
```{r fig.height=8, fig.width=14}
fisheries_long <- fisheries_total %>% 
  pivot_longer(cols = c(capture, aquaculture),
               names_to = "type",
               values_to = "tonnage") %>% 
  select(-total)

fisheries_long <- fisheries_long %>% 
  group_by(country) %>% 
  mutate(total_tonnage = sum(tonnage)) %>% 
  ungroup()

fisheries_long %>% 
  ggplot(aes(x = reorder(country, -total_tonnage), y = tonnage, fill = type)) + 
  geom_col() +
  scale_y_continuous(labels = comma) +
  scale_x_discrete(labels = ~str_wrap(.x, width = 10)) + 
  labs(title = "Tonnage of fish captured and farmed in total for each country",
       subtitle = "Capture and aquaculture",
       x = "Country",
       y = "Tonnage") +
  theme(text = element_text(size = 12.5))
```

Now, this includes more information, and it makes sense. 